Composing data visualizations

### 2020-04-26 wip-compost/paper.tex
Scale inference and example

\section{Composing visualization primitives}
I will introduce individual features of the Compost library gradually. The first important aspect of
Compost is that properties of shapes are defined in terms of domain-specific values. I first
explain what this means and then use domain-specific values to specify the core part of the
UK election results bar chart.

\subsection{Domain-specific values}

@@

\begin{figure}
\includegraphics[scale=1,trim={0cm 6.5cm 6cm 0cm},clip]{figures/values.pdf}
\caption{On a continuous scale (above), an exact position is determined by a number.
  On a categorical scale (below), an exact position is determined by the category and a
  numerical ratio from 0 to 1.}
\label{fig:scales}
\end{figure}

Figure~\ref{fig:scales} illustrates the two kinds of values using the axes from the UK
election results chart. More formally, we define a value $v$ as either a continuous value
$\kvd{cont}~n$ containing any number $n$ or a categorical value $\kvd{cat}~c, r$, consisting
of a categorical value $c$ (implemented as a string) and a ratio $r$ between $0$ and $1$:
%
\begin{equation*}
\begin{array}{rcl}
v & = & \kvd{cat}~c, r \lsep \kvd{cont}~n
\end{array}
\end{equation*}

\vspace{-2.5em}
\subsection{Visualization primitives}

\begin{equation*}
\begin{array}{rcl}
s & = & \kvd{overlay}~[\,s_1, \ldots, s_n\,]\\
 & | & \kvd{text}~\gamma, v_x, v_y, t\\
 & | & \kvd{line}~\gamma, v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\\
 & | & \kvd{fill}~\gamma, v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\\
 & | & \kvd{axis}_{x/y} s
\end{array}
\end{equation*}

xx

\begin{equation*}
\begin{array}{l}
  \kvd{axis}_x~(\kvd{axis}_y~(\kvd{overlay}~[\\
  \quad \fkvd{for}~\ident{party}, \ident{clr17}, \ident{clr19}, \ident{mp17}, \ident{mp19}~\fkvd{in}~\ident{elections}~\rightarrow\\
  \quad\quad \kvd{padding}~0,10,0,10,\,\kvd{overlay}~[\\
  \quad\quad\quad \kvd{fill}~\ident{clr},~(\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\ident{mp17}),\\
  \quad\quad\quad \hspace{2.65em}           ~(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mp17}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0})\\
  \quad\quad\quad \kvd{fill}~\ident{clr},~(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mp19}),\\
  \quad\quad\quad \hspace{2.65em}           ~(\kvd{cat}~\ident{party}, \num{1}), (\kvd{cont}~\ident{mp19}), (\kvd{cat}~\ident{party}, \num{1}), (\kvd{cont}~\num{0})~~]\\
  ]\,)\,)
\end{array}
\end{equation*}

\subsection{Inferring scales and projections}

\section{Fancy scale features}

\section{Definining abstractions}

@@

\section{Implementation structure}

@@

\section{Domain-specific language}
\label{sec:intro}
\subsection*{Domain-specific values}

\subsection*{Language of composable charts}
Now that we defined our representation of values, we can define a basic composable language of
data visualizations. A shape $s$ represents a chart. In the most basic form, it can be text label,
line connecting a list of points and a filled polygon determined by a list of points. For line,
text and polygon, we also include a parameter $\gamma$ indicating the color to be used. In addition,
a shape can be constructed by overlaying several other shapes:
s & = & \kvd{overlay}~s_1, \ldots, s_n\\
 & | & \kvd{text}~\gamma, v_x, v_y, t\\
 & | & \kvd{line}~\gamma, v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\\
 & | & \kvd{fill}~\gamma, v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}
As an example, consider bar chart with \strf{UK} and \strf{CZ} as two categorical values on the X
axis and their population in millions (continuous value) on the Y axis. This can be constructed by
creating two filled rectangles using \kvd{fill} and overlaying them using \kvd{overlay}:
\kvd{overlay}\\
\quad \kvd{fill}~\strf{\#0000ff},
 ~(\kvd{cat}~\strf{UK}, \num{0}), (\kvd{cont}~\num{00.00}), (\kvd{cat}~\strf{UK}, \num{0}), (\kvd{cont}~\num{66.04}),\\
\hspace{5.9em}(\kvd{cat}~\strf{UK}, \num{1}), (\kvd{cont}~\num{66.04}), (\kvd{cat}~\strf{UK}, \num{1}), (\kvd{cont}~\num{00.00}),\\
\quad \kvd{fill}~\strf{\#ff0000},
 ~(\kvd{cat}~\strf{CZ}, \num{0}), (\kvd{cont}~\num{00.00}), (\kvd{cat}~\strf{CZ}, \num{0}), (\kvd{cont}~\num{10.58}),\\
\hspace{5.9em}(\kvd{cat}~\strf{CZ}, \num{1}), (\kvd{cont}~\num{10.58}), (\kvd{cat}~\strf{CZ}, \num{1}), (\kvd{cont}~\num{00.00})\\
The shape specification overlays two bars of different colours. To construct a rectangle, we
define four points. For the UK, two of the points have Y value set to 0 (bottom of the bar) and
two have it set to 66.04 (top of the bar indiciating the population value). The values on the X
axis are categorical \strf{UK} values -- two are on the left as indicated by the ratio $\num{0}$
and two are on the right as indicated by ratio $\num{1}$. The ratio is always relative to the
category, so the ratios for the second bar are also $\num{0}$ and $\num{1}$.
\subsection*{Calculating scales and projections}
When processing shape specification such as the one given in the previous section, the Compost
library infers the scales for the X and Y axes and computes a projection function that can be
used for mapping domain-specific values to pixels. While doing this, it also checks for consistency --
all values on one axis have to be either categorical or continuous, but not a mix of the two.
The first step is to determine the scales for each of the axes. A scale can be continuous,
defined by a minimal and maximal value, or a categorical, defined by a list of categorical values.
Note that we do not need to track the ratios used as each category
will be allocated equal amount of space, regardless of where in that space shapes need to
appear. A scale $l$ is thus defined as:
l & = & \kvd{categorical}~c_1, \ldots, c_k \lsep \kvd{continuous}~n_{min}, \ldots, n_{max}
A scale is obtained by recursively walking over the shape and gradually constructing two scales,
for X and Y axis, from the X and Y coordinates that appear in the shape. The process is fairly
straightforward and we won't discuss it in detail (for now). The first value encountered determines
what kind of a scale will be constructed. If the type of a later value does not match the type of a
scale, the process fails.
Once we obtain scales for both X and Y axes, we calculate a projection function for each axis.
The function takes a value $v$ together with total space available (say, in pixels) and produces
a position in pixels corresponding to $v$. We also omit the details here (for now), but briefly --
for continuous values, we produce a simple linear transformation; for categorical values, we split
the available space into $k$ equally sized regions (where $k$ is the number of categorical values
in the scale) and then map a categorical value $\kvd{cat}~c, r$ to the region corresponding to $c$
according to the ratio $r$.

>>>>

\section{Composing simple visualizations}
I will introduce individual features of the Compost library gradually. The first important aspect of
Compost is that properties of shapes are defined in terms of domain-specific values. I first
explain what this means and then use domain-specific values to specify the core part of the
UK election results bar chart.

\begin{figure}[t]
\includegraphics[scale=1,trim={0cm 6.5cm 6cm 0cm},clip]{figures/values.pdf}
\caption{On a continuous scale (above), an exact position is determined by a number.
  On a categorical scale (below), an exact position is determined by the category and a
  numerical ratio from 0 to 1.}
\label{fig:scales}
\end{figure}

\subsection{Working with domain-specific values}

@@

Figure~\ref{fig:scales} illustrates the two kinds of values using the axes from the UK
election results chart. In Figure~\ref{fig:shape}, we define a value $v$ as either a continuous value
$\kvd{cont}~n$ containing any number $n$ or a categorical value $\kvd{cat}~c, r$, consisting
of a categorical value $c$ (implemented as a string) and a ratio $r$ between $0$ and $1$.
%
\begin{figure}
\begin{equation*}
\begin{array}{rcl}
v & = & \kvd{cat}~c, r \\
  & | & \kvd{cont}~n\\
  ~
\end{array}
\qquad
\begin{array}{rclcl}
s & = &\kvd{line}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] & | & \kvd{overlay}~[\,s_1, \ldots, s_n\,]\\
 & | & \kvd{fill}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] & | &\kvd{axis}_{l/r/t/b}~s\\
 & | & \kvd{text}~\gamma, v_x, v_y, t & | &\kvd{padding}~n_t,n_r,n_b,n_l,s
\end{array}
\end{equation*}
\caption{Core primitives of the Compost domain-specific language. Values $v$ are either categorical
  or continuous; a shape $s$ is then defined as a simple recursive algebraic data type.}
\label{fig:shape}
\end{figure}

\subsection{Introducing basic primitives and combinators}

Now that we know how Compost represents values, we can define the basic elements of its
domain-specific langauge. A chart is represeted by the shape $s$ defined in Figure~\ref{fig:shape}.
A primitive shape can be a text label, a line connecting a list of points or a filled polygon
defined by a list of points. The position of points is specified by X and Y coordinates,
which can be either categorical or continuous values. For text, line and polygon, we also include
a parameter $\gamma$ that specifies the element color.

Figure~\ref{fig:shape} also defines three combinators. The most important is $\kvd{overlay}$,
which overlays all shapes from a given list. When doing this, Compost automatically infers the
scales of X and Y axes and calculates suitable projections using a method discussed in the next
section. Finally, $\kvd{padding}$ adds padding around a specified shape and $\kvd{axis}$ adds
an axis showing the inferred scale on the left, right, top or bottom of a given shape.
Using those primitives, we can construct the simple UK election results bar chart in Figure~\ref{fig:simple} as follows:
%
\begin{equation*}
\begin{array}{l}
\kvd{axis}_l~(\kvd{axis}_b~(\kvd{overlay}~[\\
\quad \kvd{fill}~\strf{\#0000ff},
 ~[\;\,(\kvd{cat}~\strf{Conservative}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\strf{Conservative}, \num{0}), (\kvd{cont}~\num{365}),\\
\hspace{6.7em}\;(\kvd{cat}~\strf{Conservative}, \num{1}), (\kvd{cont}~\num{365}), (\kvd{cat}~\strf{Conservative}, \num{1}), (\kvd{cont}~\num{0})\;\,],\\
\quad \kvd{fill}~\strf{\#ff0000},
 ~[\;\,(\kvd{cat}~\strf{Labour}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\strf{Labour}, \num{0}), (\kvd{cont}~\num{202}),\\
\hspace{6.7em}\;(\kvd{cat}~\strf{Labour}, \num{1}), (\kvd{cont}~\num{202}), (\kvd{cat}~\strf{Labour}, \num{1}), (\kvd{cont}~\num{0})\,\;]~]))\\
\end{array}
\end{equation*}

\noindent
The chart specification overlays two bars of different colors and then adds axes to the bottom and
left of the chart. The two bars are filled rectangles defined using four corner points. The Y
coordinates are specified as continuous values, while the X coordinates are categorical. For
the Conservative party, two of the points have the Y coordinate set to $\kvd{cont}~\num{0}$ (bottom of the bar)
and two have the Y coordinate set to $\kvd{cont}~\num{365}$ (top of the bar). The two X coordinates
are the start and the end of the range allocated for the \strf{Conservative} category,
i.e.~$\kvd{cat}~\strf{Conservative}, \num{0}$ on the left and $\kvd{cat}~\strf{Conservative}, \num{1}$
on the right.

\begin{figure}
  \includegraphics[scale=0.57]{figures/simple}
  \vspace{0.25em}
  \caption{Two charts about the UK politics: Comparison of election results from 2017 and 2019 (left)
    and GBP-USD exchange rate with highlighted areas before and after the 23 June 2016 Brexit vote.}
  \label{fig:simple}
\end{figure}

Extending the snippet to generate a grouped bar chart that shows two results
for each party as in Figure~\ref{fig:charts} is easy. Given a party $p$, we need to
generate two rectangles, one with X coordinates $\kvd{cat}~p, 0$ and $\kvd{cat}~p, 0.5$
and the other with X coordinates $\kvd{cat}~p, 0.5$ and $\kvd{cat}~p, 1$.
In the following snippet, we use a \fkvd{for} comprehension to generate the list. All remaining
constructs are primitives of the Compost domain-specific language. Assuming \ident{elections} is a
list of election results containing a five-element tuple consisting of a party name, colors for
2017 and 2019 and results for 2017 and 2019 we create the chart using:
%
\begin{equation*}
\begin{array}{l}
  \kvd{axis}_l~(\kvd{axis}_b~(\kvd{overlay}~[\\
  \quad \fkvd{for}~\ident{party}, \ident{clr17}, \ident{clr19}, \ident{mp17}, \ident{mp19}~\fkvd{in}~\ident{elections}~\rightarrow\\
  \quad\quad \kvd{padding}~0,10,0,10,\,\kvd{overlay}~[\\
  \quad\quad\quad \kvd{fill}~\ident{clr17},~[\,(\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\ident{mp17}),\\
  \quad\quad\quad \hspace{4.08em}           ~(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mp17}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0}) \,], \\
  \quad\quad\quad \kvd{fill}~\ident{clr19},~[\,(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mp19}),\\
  \quad\quad\quad \hspace{4.08em}           ~(\kvd{cat}~\ident{party}, \num{1}), (\kvd{cont}~\ident{mp19}), (\kvd{cat}~\ident{party}, \num{1}), (\kvd{cont}~\num{0})\,]~~]~~]\,)\,)\\
\end{array}
\end{equation*}

\noindent
Aside from iterating over all available parties and splitting the bar, the example also adds padding
around the bars. A padding is specified in pixels rather than in terms of domain values. This is
sometimes preferrable over, for example, drawing a bar using a range from $\num{0.05}$ to $\num{0.5}$.
One remaining feature of the chart in Figure~\ref{fig:charts} that is still missing is the caption.
We will add this in Section~\ref{sec:abstractions}.

\subsection{Inferring scales and projections automatically}

When composing shapes using the \kvd{overlay} primitive, the user does not need to specify how
to position the child elements relatively to each other. The Compost library positions the elements
automatically. This is done in two steps. First, Compost infers the \emph{scales} for X and Y
axes. A scale represents the range of values that needs to fit in the space available for the chart.
Second, Compost calculates a \emph{projection}, a mapping from domain-specific values of the scale
to the available screen space. A scale $l$ can be either categorical or continuous:

\begin{equation*}
\begin{array}{rcl}
l & = & \kvd{continuous}~n_{min}, n_{max} \lsep \kvd{categorical}~[\,c_1, \ldots, c_k\,]
\end{array}
\end{equation*}

\vspace{-0.5em}
\noindent
A continuous scale is defined by a minimal and maximal value that need to be mapped to the
available chart space. A categorical scale is defined by a list of individual categorical values.
Note that we do not need a minimal and maximal ratios of the used categorical values as Compost
will use an equal space for each category, regardless of where in this space a shape needs to
appear.

Inferring scales is done by a simple recursive function that walks over the given shape and
constructs two scales for the X and Y axis, using the X and Y coordinates that appear in the shape.
Most of the work is done by a simple helper function that takes two scales, $l_1$ and $l_2$,
and produces a new scale that represents the union of the two:
%
\begin{equation*}
\begin{array}{l}
\ident{union}~(\kvd{continuous}~n_l, n_h)~(\kvd{continuous}~n'_l, n'_h) =\\
\qquad\kvd{continuous}~\min(n_l, n'_l), \max(n_h, n'_h)\\[0.5em]
\ident{union}~(\kvd{categorical}~[\,c_1, \ldots, c_p\,])~(\kvd{categorical}~[\,c'_1, \ldots, c'_q\,]) =\\
\qquad \kvd{categorical}~[\,c_1, \ldots, c_p\,] \;@\; [\,c'_i~|~ \forall i\in 1\ldots q, \nexists j. c_j = c'_i \;]
\end{array}
\end{equation*}

\vspace{-0.5em}
\noindent
When unioning two continuous scales, the minimum and maximum of the resulting scale is the smallest
and largest of the two minimums and maximums, respectively. When unioning two categorical scales,
we take all values of the first scale and append all values of the second scale that do not appear
in the first one. Note that this means that the order of categorical values in a scale depends on
the order in which they appear in the shape. (A possible improvement to Compost would be to support
ordinal values, which are categorical values with a well-defined ordering.) It is also worth noting
that a categorical scale cannot be combined with a continuous scale. In other words, mixing
categorical and continuous values in a single scale results in an error.

Once Compost computes scales for the given shape and all its sub-shapes, it constructs a projection
function that maps domain-specific values to the available chart space. I say more about this in
Section~\ref{sec:impl}. For a continuous scale, the projection is a linear transformation. For
categorical scale with $k$ values, we split the available chart space into $k$ equally sized
regions and then map a categorical value $\kvd{cat}~c, r$ to the region corresponding to $c$
according to the ratio $r$.

\section{Composing advanced visualizations}
\label{sec:fancy}

\subsection{Creating scales with nice ranges}

\subsection{Controlling the composition of scales}

\section{Definining abstractions}
\label{sec:abstractions}

@@

\section{Implementation structure}
\label{sec:impl}

@@
